A Computational Account of Default Mode Function\\ by Control Theory and Reinforcement Learning

### 2017-03-28 paper/paper.tex
monstrous pass some technical parts

Consider an agent faced with the choice of the next action
and being guided by rich
reenactment of really-happened, hypothetically conceived, and
predicted future events to perfect ongoing behavioral performance.
A particularly attractive computational framework
to describe, quantify, and predict autonomously acting systems like the brain
can be the combination of control theory and reinforcement learning.
It is known that, the more the external world is predictable,
the more mental activity becomes detached from the actual sensory environment
\citep{antrobus1966studies, pope1978regulation}.
Conversely,
the more the currently executed task is unknown and unpracticed,
the lower the tendency for stimulus-independent thoughts
\citep{filler1973daydreaming, teasdale1995stimulus}.
These ``offline'' processes may however contribute to optimized control of the organism.
% We formalize
% a policy matrix to capture the space of possible actions that can be performed
% on the environment given the current state. A value matrix
% maps expected rewards to environmental objects and events.
% Switching between states reduces to a sequential processing model.
% In a model-based reinforcement learning (RL) approach,
% we try to understand the environment always better
% starting, at birth, with an environment that is completely unknown.
Informed by outcomes of performed action,
the DMN dynamics are constantly adapted in feedback loops
that are calibrated by prediction error.
A DMN framework incorporating RL can naturally embed human behavior
into the tension between exploitative action with immediate gains and
exploratory action with longer-term reward schedules.

@@

\subsection{Value functions and policies}
The behavior of the agent is governed by some kind of \textit{policy},
which maps states of the world to
a set of candidate actions to perform given this state. Starting a time $t=0$,
a policy $\pi$ generates a trajectory as follows
\begin{eqnarray*}
  \begin{split}
    S_1 &\sim p(.|S_0,A_0), \text{ collect rewards }R_0 = r(S_0, A_0)\\
    A_1 &\sim \pi(.|S_1)\\
    S_2 &\sim p(.|S_1,A_1), \text{ collect rewards }R_1 = r(S_1, A_1)\\
    \ldots\\
    A_{t} &\sim \pi(.|S_{t})\\
    S_{t+1} &\sim p(.|S_{t},A_{t}), \text{ collect rewards }R_{t} = r(S_{t}, A_{t})\\
    \ldots
  \end{split}
\end{eqnarray*}
Since an action taken now might have repercussions long into the feature, it turns out that the
quantity to optimize is not the instantaneous rewards $r(s, a)$, but a
cumulative reward estimate which takes into account the future. A standard for
modeling this accumulation is the so-called
time-discounted \textit{cumulative reward} function
\begin{equation}
  \label{eq:cumr}
  G^\pi = \sum_{t=0}^{\infty}\gamma^{t}R_t.
\end{equation}
This random variable\footnote{Random as it depends both on the environment's dynamic and the
  policy $\pi$ being played (which can be stochastic).}  measures the cumulative reward of
following a policy $\pi$.
\begin{mdframed}
  \textbf{Where is ``value'' buffered in the brain ?}
  The amygdala is known to be involved in significance evaluation) and
  the nucleus accumbens (NAc) is involved in reward evaluation.
\end{mdframed}
  The  goal of RL is then adapt this policy so as to  maximize the cumulative rewards. In
  the definition of $G^\pi$ above, the constant $\gamma$ $(0 \le \gamma < 1$) is the reward \textit{discount factor}.
Taking $\gamma = 0$   corresponds to a perfectly myopic agent who is solely concerned about their
immediate rewards and has no horizon for the future, which is not compatible with long-term goal planning
as implemented in the DMN. For there to be any learning at all, it is thus necessary that $0 < \gamma < 1$.
% \footnote{In which case we can take $T \rightarrow +\infty$, provided the reward function $r$ is bounded.}.
Thus the agent is positively biased towards considering rewards which are will
come soon. The reward importance hence decays exponentially
with distance in the future,
given that there is more uncertainty in the farsighted future.
The reward discount factor $\gamma$ can thus be seen as calibrating
the delay discounting behavior of the intelligent agent.
It is also important to appreciate that a stochastic policy estimation
is more advantageous in many RL settings.
For instance, a deterministic strategy in playing rock-paper-scissors can
be easily exploited and a uniformly random choice is optimal.

The agent's goal can then be formulated as finding policy $\pi$ which maximizes the
expected cumulative value of state-stat-pairs $(A_0,S_0)$, also known as the \textit{Q-value}, namely
\begin{equation}
  \label{eq:qval}
  Q^{\pi}(s,a) = \mathbb E [G^\pi|S_0=s,A_0=a,A_1 \sim \pi,A_2 \sim \pi,\ldots].
\end{equation}
\begin{mdframed}
In plain English, the Q-value $Q^{\pi}(s,a)$ is the expected / average / mean future reward
over all possible trajectories in which the agent starts off starts at with an environment in state
$s$, takes and action $a$, and then follows the policy $\pi$ thereafter to select future actions.
\end{mdframed}
%% Given that each $\pi$ correspondence to an exhaustive description of
%% behavioral patterns, $Q^{\pi}(S_t,A_t)$ corresponds to the epistemic value of
%% the set of actions given states in finite time.
By construction, maximizing the expectation in  \eqref{eq:qval} is intractable in general
due to the enormous number of possible trajectories the agent might need to
consider.
% \begin{mdframed}
%   \textbf{N.B.:}
%   If the Bellman-based computational ideas have a corresponding
%   neurobiological implementation, this provides a framework of how
%   optimal behavior could be computable in the human brain.
% \end{mdframed}
One popular simplification that removes this issue is to only consider the sub-family of
\textit{deterministic policies}, which map each state to the best single best action which
should be taken at thin state, and therefore define functions from states to actions.
\textit{Q-learning} (Watkins \& Dayan, 1992), a commonly used off-policy algorithm, uses the
\textit{greedy policy}
\begin{equation}
  \pi(s) = \argmax_{a \in \mathcal A}{Q}(s, a).
  \label{eq:qlearning}
\end{equation}
% where $\tilde{Q}(s, a)$ is the current estimate of the state-action value
% table.
The celebrated \textit{Bellman's equation} \cite{sutton1998reinforcement} then takes the simple form
\begin{equation}
  \begin{split}
    Q^\pi(s, a) &= \mathbb E_{s' \sim p(.|s,a)} [r(s,a) + \gamma \max_{a' \in \mathcal A}Q^\pi(s', a')]
    \\
    &= \mathbb E_{s' \sim p(.|s,a)} [r(s,a) + \gamma Q^\pi(s', \pi(s'))],
  \end{split}
  \label{eq:bellman}
\end{equation}
a \textit{fixed-point} equation which provides a recursive decomposition of optimal behavior.
A complicated dynamic programming optimization can be
broken into simpler sub-problems at different time points,
which has often been closely related to the medial prefrontal part of the DMN
\citep{koechlin1999role, braver2002role}.
Using the Bellman equation, each state can be associated with a certain value,
to guide action towards a better state, thus improving on the agent's current policy.
% It is a fixed-point equation for the deterministic policy $\pi$.
Notice that in the expectation in \eqref{eq:bellman}, the sampling is done over only things which
depend on the environment, and so can be learned off-policy by observing state-transitions
triggered by another behavioral policy, which can be stochastic.

\subsection{Efficient RL with value approximation and replay}
%As already mentioned in the previous section, Q-learning optimizes over the class of
% deterministic policies of the form \eqref{eq:qlearning}.
A successful RL theory needs ways to approximate the state-action value function
which scale up to large state spaces and / or action spaces.
In state $s \in \mathcal S$, the agent takes an action $a \in \mathcal A$
% (for example, $a$ = ``Read'', or ``Run'', or ``Laugh'', or ``Sympathize'', or ``Empathize'', etc.)
by sampling from its current policy matrix, and collects a reward $r$,
% (measured on their internal currency of worldly values ?),
and the environment transitions to a new state $s' \in \mathcal S$. At the end of
this step, a new \textit{experience} $e = (s,a,r,s')$ is produced; this represents an exemplar
behavior of the agent and is recorded in replay memory buffer:
$D \leftarrow D \cup \{e\}$ (possibly discarding the oldest entries to make space).
Now, at iteration $k+1$, replay consists in sampling (uniform or importance-weighted
\footnote{e.g weighted by TD error of the state transition $s \overset{a}{\rightarrow} s'$.} ?)  mini-batches of experiences
$(s, a, r, s') \sim \mathcal U(D)$ from the replay memory $\mathcal D$ (hyppocampus ?)
and trying to approximate
the would-be $Q$-value for the state-action pair $(s,a)$ as predicted by Bellman's equation \eqref{eq:bellman}, namely
$r + \gamma \max_{a'} Q(s', a'|\theta^Q_k)$, with the output of a parametrized regression model $(s,a,r,s')
\mapsto {Q}(s, a|\theta^Q_{k})$, where $Q(.,.|\theta^Q_{k})$ is an approximator for the $Q$-value operator,  parameterized by $\theta^Q_{k+1}$.
\begin{mdframed}
In the brain, such an experience replay mechanism is probably implemented in the hyppocampus (HC). Indeed,
hippocampal replay is a phenomenon observed in rats, mice, cats, rabbits, songbirds, and monkeys. Indeed, during sleep or awake rest, replay refers to the re-occurrence of a sequence of cell activations that also occurred during activity, but the replay has a much faster time scale.
\end{mdframed}

@@

\paragraph{Link to classical RL algorithms.}
Note that most RL algorithms, including \textit{Temporal Difference (TD)}
learning \cite{sutton1998reinforcement}, REINFORCE \cite{williams1992}, and SARSA can be
cast in this general variational framework. For example TD corresponds to
the above framework using a linear value approximator with feature encoding
$\phi(s,a) = \delta_{(s,a)} =  \text{ point mass at }(s,a)$ on the grid
$\mathcal S \times \mathcal A$, and so
$$\nabla_{\theta^Q_{k+1}}Q(s, a|\theta^Q_{k+1}) = \phi(s, a) = \delta_{(s,a)}.$$
%% Whence, the gradient update due to the sample $(s,a,r,s') \in \mathcal D$ is
%% $$\theta^Q(s,a) \leftarrow \theta^Q(s,a) + (Q(s, a|\theta^Q_{k+1}) - y)$$ ??

\paragraph{TODO: Monte Carlo optimization.}
Some text snippets that might be pertinent:
Monte-Carlo simulation can be used to update the agent’s belief state
\citep{silver2010monte}.
Monte-Carlo simulation provides a simple method for evaluating
a state s.
In any MDP, there is at least one optimal policy matrix that
achieves the optimal value function???
The simulator is used to generate sequences of states, observations and rewards. These simulations are used to update the value function, without ever looking inside the black box describing the model’s dynamics. In addition, Monte-Carlo methods have a sample complexity that is determined only by the underlying difficulty of the POMDP, rather than the size of the state space or observation space.
Monte-Carlo simulation can be turned into a simple control algorithm by evaluating all legal actions and selecting the action with highest evaluation [15]. Monte-Carlo simulation can be extended to partially observable MDPs.
In these challenging POMDPs, Monte-Carlo simulation provides an effective mechanism both for tree search and for belief state updates, breaking the curse of dimensionality and allowing much greater scalability than has previously been possible.
Link this to experience sampling in the hippocampus.
Much recent work points to Monte Carlo or stochastic sampling– based approximations as a unifying framework for understanding how Bayesian inference may work practically across all these levels, in minds, brains, and machines.
For sampling-based methods it is still almost non-existent beyond the description of receptive fields.

\subsection{Correspondence between DMN function and RL framework}
\begin{figure}[!h]
  \includegraphics[width=.9\linewidth]{rl_process_chart.pdf}
  \caption{The DMN as an RL agent...}
  \label{fig:rl_process_chart}
\end{figure}

\begin{itemize}
  \item The policy matrix represents the repertoire of possible actions
  on the world given a current state. It encodes the probabilites of
  choosing actions to be executed in a certain situation.
  \item Interactions between external and internal sources of
  information are probably commonplace during perception \citep{gershman2010learning}
  \item The MDP frameworks how the human brain could solve
  environmental problems over extended time periods.
  \item Given a state, the value function represents the estimated utility of
  being in that state, which is subjective to the agent.
  For a state $s$, the value function at $s$ namely $V(s)$ define provides a quantitative answer to the question:
  "How desirable is it to be in state $s$ ?"
  For each action $a$ that can be taken from state $s$,
  the Q-value  $Q(s, a)$ defined in \eqref{eq:qval} provides the subjective
  utility of executing a specific action; it answers the question
  "What is the expected utiltiy of taking action $a$ in this situation ?"
  \item The DMN may subserve
  constant exploration of possible future actions and their
  cumulative outcomes. Implicit computation of future choices
  provides an explanation for purposeful mind-wandering.
  \item hippocampus -> provides perturbed action-transition-state-reward samples
  as a block of "imagined", "hypothesized", "recalled" experience;
  the small variations in these experience blocks allow searching
  a larger space of parameters and possible experiences.
  In the absence of environmental input and feedback
  (e.g., mind-wandering or sleep) pseudo-experiencing possible
  future scenarios and action outcomes.
  Our approach acknowledges the unavoidable stochasticity of
  computation in neuronal processes \citep{faisal2008noise}.
  \item
  \textit{Inference} in the human brain reduces to generalization of
  policy and value adaptations from sampled experiences to
  successful action choices and reward predictions in future states.
  \item In the RL view, plasticity in the DMN arises naturally.
  If an agent behaving optimally in a given environment moves
  to novel, yet unexperienced environments, reward prediction errors will
  massively increase. This will lead to adaptation of the policy matrix
  and value function until the system converges to a
  new steady-state of optimal action decisions given certain states.
  \item The DMN is today known to consistently increase in neural
  activity when engaging cognitive processes that are detached from
  current sensory environment. DMN involvement in thinking about the past,
  the future, and hypothetical possibilities ties in with the
  MDP's implicit computation of action and state cascades as a function
  of what happened in the past.
\end{itemize}

\paragraph*{Some notes:}
\begin{itemize}
  \item Actions are continuous, whilst states are continuous.
  \item The brain does backprop (see Hinton's recent talk at Stanford)
  \item model weights $\theta^Q$ should correspond to connections between neurons
\item humm, looks like such a knowledge would simply translate into constraints on the DqN weights $\theta^Q$ (... at least, part of such knowledge)
  ok, gotta think about the model-free / model-based reconsiliation
\end{itemize}

\subsection{Hippocampal replay = inverse reinforcement-learning ?}
%\begin{mdframed}
  Given the trace $s_0,a_0,s_1,\ldots,$ of an optimal agent's strategy $\pi^*$ in
  a POMDP (called a \textit{teacher's demonstration}), can we figure out what is the
  (instantaneous) reward function $r: \mathcal S \times \mathcal \mathcal A
  \rightarrow \mathbb R$ that the agent is optimizing over a prescribed class of reward functions
  (e.g linear rewards $r(s,a) \equiv \theta^T\phi(s,a)$)?
  For example, given traces of motor actions from a an adult to grab a cup on a table,
  can an observing child figure out what ``langragian'' functional is being minimized by the former ?
  How ? Can they reproduce this optimal behavior ?
%\end{mdframed}
Such questions are of course pertinent to our decision-making theory for the DMN.
In the general artificial case, the problem has been extensively studied and partially solved by
\cite{abbeel2004}. They are also been rigourously studied in general optimal control literature under the name
``inverse optimal control'', but in model-based certain (where the physical dynamics are known, etc.)...

@@

\section{Free-energy principles!}
The so-called free-energy principle in its present form (including notions like ``generative density'', ``recognition density'', etc.) can be traced back to works of Dayan \& Hinton \citep{dayan1995helmholtz} in which they introduced the so-called \textit{Helmholtz machine}, a hierarchical factorial directional deep belief-net (DBN).
In this subsection, we will develop from first-principles, the bare-bones minimalistic ideas needed to build a free-energy principle for general decision-making. This ideas were first developed by Hinton et al. in the early 90s in building their Helmholtz machine. Theories like Friston's free-energy principle and active-inference will then emerge as particular instances of this general framework, with particular design choises. For example the Friston theory axiomatizes that the brain uses a (problematic, as it implicitly assumes that posterior of each hidden unit is factorial, e.g) wake-sleep algorithm to train the underlying Helmholtz machine, etc.

>>>>

Consider an agent faced with the choice of the next action
and being guided by rich
reenactment of really-happened, hypothetically conceived, and
predicted future events to perfect ongoing behavioral performance.
A particularly attractive computational framework
to describe, quantify, and predict autonomously acting systems like the brain
can be the combination of control theory and reinforcement learning.
It is known that, the more the external world is predictable,
the more mental activity becomes detached from the actual sensory environment
\citep{antrobus1966studies, pope1978regulation}.
Conversely,
the more the currently executed task is unknown and unpracticed,
the lower the tendency for stimulus-independent thoughts
\citep{filler1973daydreaming, teasdale1995stimulus}.
These ``offline'' processes may however contribute to optimized control of the organism.
% We formalize
% a policy matrix to capture the space of possible actions that can be performed
% on the environment given the current state. A value matrix
% maps expected rewards to environmental objects and events.
% Switching between states reduces to a sequential processing model.
% In a model-based reinforcement learning (RL) approach,
% we try to understand the environment always better
% starting, at birth, with an environment that is completely unknown.
Informed by outcomes of performed action,
the DMN dynamics are constantly adapted in feedback loops
that are calibrated by prediction error.

A DMN framework incorporating RL can naturally embed human behavior
into the tension between exploitative action with immediate gains and
exploratory action with longer-term reward schedules.

@@

\subsection{Long-term rewards and policies}
The behavior of the agent is governed by some kind of \textit{policy},
which maps states of the world to
a set of candidate actions to perform given this state. Starting a time $t=0$,
a policy $\pi$ generates a trajectory as follows
\begin{eqnarray*}
  \begin{split}
    \textbf{ observe transition: }s_1 &\sim p(s|s_0,a_0)\textbf{ and collect rewards }R_0 = r(s_0, a_0)\\
    \textbf{choose action: }a_1 &\sim \pi(a|s_1)\\
    \textbf{ observe transition: }s_2 &\sim p(s|s_1,a_1), \textbf{ and collect rewards }R_1 = r(s_1, a_1)\\
    \ldots\\
    \textbf{choose action: }a_{t} &\sim \pi(a|s_{t})\\
    \textbf{ observe transition: }s_{t+1} &\sim p(s|s_{t},a_{t}), \textbf{ and collect rewards }R_{t} = r(s_{t}, a_{t})\\
    \ldots
  \end{split}
\end{eqnarray*}
Since an action taken now might have repercussions long into the feature, it turns out that the
quantity to optimize is not the instantaneous rewards $r(s, a)$, but a
cumulative reward estimate which takes into account the future. A standard for
modeling this accumulation is the so-called
time-discounted \textit{cumulative reward} function
\begin{equation}
  \label{eq:cumr}
  G^\pi = \sum_{t=0}^{\infty}\gamma^{t}R_t.
\end{equation}
This random variable\footnote{Random as it depends both on the environment's dynamic and the
  policy $\pi$ being played (which can be stochastic).}  measures the cumulative reward of
following a policy $\pi$.
\begin{mdframed}
  \textbf{Where is ``value'' buffered in the brain ?}
  The amygdala is known to be involved in significance evaluation) and
  the nucleus accumbens (NAc) is involved in reward evaluation.
\end{mdframed}
  The  goal of RL is then adapt this policy so as to  maximize the cumulative rewards. In
  the definition of $G^\pi$ above, the constant $\gamma$ $(0 \le \gamma < 1$) is the reward \textit{discount factor}.
Taking $\gamma = 0$   corresponds to a perfectly myopic agent who is solely concerned about their
immediate rewards and has no horizon for the future, which is not compatible with long-term goal planning
as implemented in the DMN. For there to be any learning at all, it is thus necessary that $0 < \gamma < 1$.
% \footnote{In which case we can take $T \rightarrow +\infty$, provided the reward function $r$ is bounded.}.
Thus the agent is positively biased towards considering rewards which are will
come soon. The reward importance hence decays exponentially
with distance in the future,
given that there is more uncertainty in the farsighted future.
The reward discount factor $\gamma$ can thus be seen as calibrating
the delay discounting behavior of the intelligent agent.
It is also important to appreciate that a stochastic policy estimation
is more advantageous in many RL settings.
For instance, a deterministic strategy in playing rock-paper-scissors can
be easily exploited and a uniformly random choice is optimal.

\subsection{The Q-value: a goodness measure for actions}
The agent's goal can then be formulated as finding policy $\pi$ which maximizes the
expected cumulative value of an state-action pair $(a,s)$, also known as the \textit{Q-value}, namely
\begin{equation}
  \label{eq:qval}
  Q^{\pi}(s,a) = \mathbb E [G^\pi|s_0=s,a_0=a,a_1 \sim \pi,a_2 \sim \pi,\ldots].
\end{equation}
In plain English, the Q-value $Q^{\pi}(s,a)$ is the expected / average / mean future reward
over all possible trajectories in which the agent starts off starts at with an environment in state
$s$, takes and action $a$, and then follows the policy $\pi$ thereafter to select future actions.
%% Given that each $\pi$ correspondence to an exhaustive description of
%% behavioral patterns, $Q^{\pi}(s_t,a_t)$ corresponds to the epistemic value of
%% the set of actions given states in finite time.
For the brain,
$Q^{\pi}(s, a)$ defined in \eqref{eq:qval} provides the subjective
utility of executing a specific action; it answers the question
"What is the expected utiltiy of taking action $a$ in this situation ?".
When a choice of actions is available, the $Q^{\pi}(s,a)$
provides a formalization of optimal selection.

\subsection{Computing optimal behavior}
By construction, maximizing the expectation in  \eqref{eq:qval} is intractable in general
due to the enormous number of possible trajectories the agent might need to
consider.
% \begin{mdframed}
%   \textbf{N.B.:}
%   If the Bellman-based computational ideas have a corresponding
%   neurobiological implementation, this provides a framework of how
%   optimal behavior could be computable in the human brain.
% \end{mdframed}
One popular simplification that removes this issue is to only consider the sub-family of
\textit{deterministic policies}, which map each state to the best single best action which
should be taken at thin state, and therefore define functions from states to actions.

\subsubsection{Q-learning} Q-learning (Watkins \& Dayan, 1992), a commonly used off-policy algorithm, uses the
\textit{greedy policy}
\begin{equation}
  \pi(s) = \argmax_{a \in \mathcal A}{Q}(s, a).
  \label{eq:qlearning}
\end{equation}
% where $\tilde{Q}(s, a)$ is the current estimate of the state-action value
% table.
The celebrated \textit{Bellman's equation} \cite{sutton1998reinforcement} then takes the simple form
\begin{equation}
  \begin{split}
    Q^\pi(s, a) &= \mathbb E_{s' \sim p(s'|s,a)} [r(s,a) + \gamma \max_{a' \in \mathcal A}Q^\pi(s', a')]
    \\
    &= \mathbb E_{s' \sim p(s'|s,a)} [r(s,a) + \gamma Q^\pi(s', \pi(s'))],
  \end{split}
  \label{eq:bellman}
\end{equation}
a \textit{fixed-point} equation which provides a recursive decomposition of optimal behavior.
A complicated dynamic programming optimization can be
broken into simpler sub-problems at different time points,
which has often been closely related to the medial prefrontal part of the DMN
\citep{koechlin1999role, braver2002role}.
Using the Bellman equation, each state can be associated with a certain value,
to guide action towards a better state, thus improving on the agent's current policy.
% It is a fixed-point equation for the deterministic policy $\pi$.
Notice that in the expectation in \eqref{eq:bellman}, the sampling is done over only things which
depend on the environment, and so can be learned off-policy by observing state-transitions
triggered by another behavioral policy, which can be stochastic.

\subsubsection{Efficient learning via value approximation and experience replay}
%As already mentioned in the previous section, Q-learning optimizes over the class of
% deterministic policies of the form \eqref{eq:qlearning}.
A successful RL theory needs ways to approximate the state-action value function
which scale up to large state spaces and / or action spaces.
In state $s \in \mathcal S$, the agent takes an action $a \in \mathcal A$
% (for example, $a$ = ``Read'', or ``Run'', or ``Laugh'', or ``Sympathize'', or ``Empathize'', etc.)
by sampling from its current policy matrix, and collects a reward $r$,
% (measured on their internal currency of worldly values ?),
and the environment transitions to a new state $s' \in \mathcal S$. At the end of
this step, a new \textit{experience} $e = (s,a,r,s')$ is produced; this represents an exemplar
behavior of the agent and is recorded in replay memory buffer:
$D \leftarrow D \cup \{e\}$ (possibly discarding the oldest entries to make space).
Now, at iteration $k+1$, replay consists in sampling (uniform or importance-weighted
\footnote{e.g weighted by TD error of the state transition $s \overset{a}{\rightarrow} s'$.} ?)  mini-batches of experiences
$(s, a, r, s') \sim \mathcal U(D)$ from the replay memory $\mathcal D$ (hyppocampus ?)
and trying to approximate
the would-be $Q$-value for the state-action pair $(s,a)$ as predicted by Bellman's equation \eqref{eq:bellman}, namely
$r + \gamma \max_{a'} Q(s', a'|\theta^Q_k)$, with the output of a parametrized regression model $(s,a,r,s')
\mapsto {Q}(s, a|\theta^Q_{k})$, where $Q(.,.|\theta^Q_{k})$ is an approximator for the $Q$-value operator,  parameterized by $\theta^Q_{k+1}$.

In the brain, such an experience replay mechanism is probably implemented in the hyppocampus (HC). Indeed,
hippocampal replay is a phenomenon observed in rats, mice, cats, rabbits, songbirds, and monkeys. Indeed, during sleep or awake rest, replay refers to the re-occurrence of a sequence of cell activations that also occurred during activity, but the replay has a much faster time scale.

@@

\begin{mdframed}
\textbf{A note on classical RL algorithms.}
Note that most RL algorithms, including \textit{Temporal Difference (TD)}
learning \cite{sutton1998reinforcement}, REINFORCE \cite{williams1992}, and SARSA can be
cast in this general variational framework. For example TD corresponds to
the above framework using a linear value approximator with feature encoding
$\phi(s,a) = \delta_{(s,a)} =  \text{ point mass at }(s,a)$ on the grid
$\mathcal S \times \mathcal A$, and so
$$\nabla_{\theta^Q_{k+1}}Q(s, a|\theta^Q_{k+1}) = \phi(s, a) = \delta_{(s,a)}.$$
%% Whence, the gradient update due to the sample $(s,a,r,s') \in \mathcal D$ is
%% $$\theta^Q(s,a) \leftarrow \theta^Q(s,a) + (Q(s, a|\theta^Q_{k+1}) - y)$$ ??
\end{mdframed}

\subsubsection{Interlude: zoom on the Hippocampus}
\paragraph{Does the hippocampus do Monte-Carlo sampling ?}
In RL, Monte-Carlo simulation can be used to update the agent's belief state
\citep{silver2010monte}. Such methods have a sample complexity that is determined only by the underlying difficulty of the POMDP, rather than the size of the state space or observation space,
which can be prohibitively large.
Monte-Carlo simulation provides a simple method for evaluating a state $s$.
They provide an effective mechanism both for tree search and for belief state updates, breaking the curse of dimensionality and allowing much greater scalability than has previously been possible.

In the brain, the HC would generate / imagine sequences of world states, observations and rewards. These simulations would be used to update the value function, without ever looking inside the black box describing the model's dynamics. This would be a simple control algorithm by evaluating all legal actions and selecting the action with highest evaluation.
In these challenging POMDPs, Monte-Carlo simulation provides an effective mechanism both for tree search and for belief state updates, breaking the curse of dimensionality and allowing much greater scalability than has previously been possible.
Much recent work points to Monte Carlo or stochastic sampling– based approximations as a unifying framework for understanding how Bayesian inference may work practically across all these levels, in minds, brains, and machines. For sampling-based methods it is still almost non-existent beyond the description of receptive fields.

\paragraph{Hippocampal replay = inverse reinforcement-learning ?}
%\begin{mdframed}
  Given the trace $s_0,a_0,s_1,\ldots,$ of an optimal agent's strategy $\pi^*$ in
  a POMDP (called a \textit{teacher's demonstration}), can we figure out what is the
  (instantaneous) reward function $r: \mathcal S \times \mathcal \mathcal A
  \rightarrow \mathbb R$ that the agent is optimizing over a prescribed class of reward functions
  (e.g linear rewards $r(s,a) \equiv \theta^T\phi(s,a)$)?
  For example, given traces of motor actions from a an adult to grab a cup on a table,
  can an observing child figure out what ``langragian'' functional is being minimized by the former ?
  How ? Can they reproduce this optimal behavior ?
%\end{mdframed}
Such questions are of course pertinent to our decision-making theory for the DMN.
In the general artificial case, the problem has been extensively studied and partially solved by
\cite{abbeel2004}. They are also been rigourously studied in general optimal control literature under the name
``inverse optimal control'', but in model-based certain (where the physical dynamics are known, etc.)...

@@

\subsection{Conclusion: putting everything together}
\begin{figure}[!h]
  \includegraphics[width=.9\linewidth]{rl_process_chart.pdf}
  \caption{The DMN as an RL agent...}
  \label{fig:rl_process_chart}
\end{figure}
The DMN is today known to consistently increase in neural
  activity when engaging cognitive processes that are detached from
  current sensory environment \citep{kenet2003spontaneously, fiser2004small}.
  Its involvement in thinking about the past,
  the future, and hypothetical possibilities ties in with the implicit computation of
  action and state cascades as a function of what happened in the past.
% A policy represents the repertoire of possible actions
%   on the world given a current state. It encodes the probabilites of
%   choosing actions to be executed in a certain situation.
% Interactions between external and internal sources of
%   information are probably commonplace during perception \citep{gershman2010learning}
The DMN may subserve
  constant exploration of possible future actions and their
  cumulative outcomes. Implicit computation of future choices
  provides an explanation for purposeful mind-wandering.

The HC provides perturbed action-transition-state-reward samples
  as a block of "imagined", "hypothesized", "recalled" experience;
  the small variations in these experience blocks allow searching
  a larger space of parameters and possible experiences.
  In the absence of environmental input and feedback
  (e.g., mind-wandering or sleep) pseudo-experiencing possible
  future scenarios and action outcomes.
  Our approach acknowledges the unavoidable stochasticity of
  computation in neuronal processes \citep{faisal2008noise}.

In our proposed RL-based view, \textit{inference} in the human brain reduces to generalization of
  policy and value adaptations from sampled experiences to
  successful action choices and reward predictions in future states.
  Plasticity in the DMN also arises naturally:
  If an agent behaving optimally in a given environment moves
  to novel, yet unexperienced environments, reward prediction errors will
  massively increase. This will lead to adaptation of the policy matrix
  and value function until the system converges to a
  new steady-state of optimal action decisions given certain states.

% \paragraph*{Some notes:}
% \begin{itemize}
%   \item Actions are continuous, whilst states are continuous.
%   \item The brain does backprop (see Hinton's recent talk at Stanford)
%   \item model weights $\theta^Q$ should correspond to connections between neurons
% \item humm, looks like such a knowledge would simply translate into constraints on the DqN weights $\theta^Q$ (... at least, part of such knowledge)
%   ok, gotta think about the model-free / model-based reconsiliation
% \end{itemize}

\section{Relation to other models decision-making and optimal control for the brain}
\subsection{The free-energy principle and active inference}
% \subsubsection{The theory}
In Friston's free-energy principle (FEP) \citep{friston2010free,fristonAIorRL}, the brain emerges as biomechanical inference engine which must minimize the long-term average of surprise.
Precursors of this theory can be traced back to \citep{dayan1995helmholtz} in which they introduced the so-called \textit{Helmholtz machine}, a hierarchical factorial \textit{directional deep belief-net (DBN)}. According to FEP's account, the goal of the brain is then to optimize over the generative model $G$: to iteratively
modify its internal representation $p_G(\z|\x)$ about objects in the world, their interactions and dynamics, etc., so as to minimize surprise when these representations are confronted with sensory input $\x$ during perception cycles. These is called the \textit{generative} model. FEP also postulates a dual model called the \textit{recognition} model,
which works in tandem with the generative model $p_R(\z|\x)$, to accomplish approximate inference. The recognition model dreams / fantasizes imaginary worlds $\z$ whilst the latter tries to generate sensations $\x$ which match these dreams!

Because surprisal is intractably difficult to optimize (since we would need to sum over all hidden causes of the sensations), FEP sets out to instead minimize an upper-bound thereupon, namely the \text{free-energy} $F_G^R(\x)$ given by

  \begin{equation}
    \begin{split}
      \text{generative surprise } &:= -\log(p_G(\x)) = F_G(\x) \\
      &=\underbrace{F^R_G(\x)}_{\text{accuracy}} - \underbrace{D_{KL}(P_R(.|\x) || P_G(.|\x))}_{\text{complexity}} \\
      &\le F^R_G(\x),
      \text{ with equality if }p_R(\z|\x) = p_G(\z|\x)\text{ for all } \z
    \end{split}
  \end{equation}

% \begin{figure}[!tb]
%   \centering
%   \includegraphics[width=.8\linewidth]{vae.pdf}
%   \caption{Variational autoencoders...}
%   \label{fig:vae}
% \end{figure}
The main algorithm for minimizing free-energy $F_G^R(\x)$ is the \textit{wake-sleep algorithm}.
\citep{dayan1995helmholtz}. As noted already in \citep{dayan1995helmholtz},
a crucial drawback of the wake-sleep algorithm is that it requires a concurrent models (generative and recogntiion), which together do not correspond to optimization of (a bound of) the marginal likelihood (because of the incorrect KL used therein, etc.). Thus the brain could not possible be running such a algorithm, not even in principle! To the rescue, we note that the recent theory of
\textit{variational auto-encoders (VAEs)} \citep{kingma2013auto} might provide an efficient alternative to the wake-sleep algorithm, as it overcomes the technical limits of the former, by using a reparametrization trick. For example, unlike the wake-sleep algorithm for minimizing free-energy, VAEs can be efficiently trained back-propagation of learning errors.

\subsubsection{Comparison to our proposed theory}
On the surface, a common point between the FEP and our proposed RL-based framework
  is placing the minimization of a surprise signal at the core of brain function.
  Indeed in RL, surprise minimization is subsumed  by accurate prediction of
  rewarding outcomes in the future. A ``free-energy'' agent is barely a biomechanical machine which has the tendency to resist internal phase-transitions which lead contradictory with the
  blurred version of the external world captured as sensations
  \citep{friston2010free,fristonAIorRL,ortega2013thermodynamics}. Such a theory cannot by itself,
  explain the emergence of strategic behavior inherent in humans (e.g cite dark-room experiment).

% Friston's critic of RL \citep{fristonAIorRL} (he proposes ``active inference''): "This equation (equation 18) comes from the theory of dynamic programming,
% pioneered in the 1950s by Richard Bellman and colleagues [2]. To
% optimise control a~pð Þ x~ under this formulation, we have to: (i)
% assume the hidden states are available to the agent and (ii) solve
% Equation 18 for the value-function. Solving for the value-function
% is a non-trivial problem and usually involves backwards induction
% or some approximation scheme like reinforcement learning [4–6].
% The free-energy formulation circumvents these problems by
% prescribing the policy in terms of free-energy, which encodes
% optimal control (Equation 6)"

% The following critics can be made:
% \begin{itemize}
%   \item
%     As noted by Dayan et al. (\textit{Variants of Helmholtz machines}), the inter-neuronal intra-layer independence assumption which is at the center of the HM becomes severely problematic as it is agnostic to the known organization of cortical layers...
%   \item A drawback of the wake-sleep algorithm is that it requires a concurrent models (generative and recognitiion), which together do not correspond to optimization of (a bound of) the marginal likelihood (because of the incorrect KL used therein, etc.).
%   \item Also, note that the wake-sleep algorithm doesn't do backprop! This is due to
%     technical difficulty in getting derivatives of loss function w.r.t
%     recognition weights $\W^R$).
%   \item This difficulty was removed in the 2010s by \citep{kingma2013auto}, an other groups, via a ``reparametrization trick''.
%   \end{itemize}

\textbf{XXX: add tons of more examples and connnections}
\textbf{XXX: integrate aspects from section "ThE MARkOv DECISION PROBLEM" in Dayan and Daw}
\textbf{XXX: integrate aspects from WakeNsleep by Hinton}
\textbf{XXX: integrate aspects from Friston2014(Dropbox) on active inference}
\textbf{XXX: integrate aspects from Sutton/Barto book chapter}
